---
title: "Regression Models on mtcars dataset"
author: "Mark Evans"
output: pdf_document
---

## Introduction

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions: 

1. "Is an automatic or manual transmission better for MPG" 
2. "Quantify the MPG difference between automatic and manual transmissions" 

## Exploratory Data Analysis

```{r}
library(datasets)
data(mtcars)
```

### 1. Is an automatic or manial transmission better for MPG?

Print the summary statistics for the mpg columns for automatic and manual.

``` {r, echo=TRUE}
summary(mtcars[mtcars$am==0, 1])
summary(mtcars[mtcars$am==1, 1])
```

Hypothesis Test with the following parameters was tested:

- Null hypothesis -  Ho: mean mpg of automatic = mean mpg of manual 
- Alternate hypothesis - Ha: mean mpg of automatic != mean mpg of manual

``` {r, echo=TRUE}
t.test(mtcars$mpg~mtcars$am)
```

Conclusion: Based on the summary above, we can reject the null hypothesis that the means of mpg between automatic and manual vehicles a are the same. We can conclude that manual transmission is better for MPG than automatic transmission, assuming all other variables remain unchanged. 

### 2. Quantify the MPG difference between automatic and manual transmissions. 

We will run 3 linear models and compare using the anova function. Print the summary of the best model at the end. 

1. mpg as the response variables and am as the predictor variable. 
2. mpg as the response variable and wt, cyl, qsec, hp, and am as predictor variables. 
3. mpg as the response variable the remainder of the dataset as predictor variables 

``` {r, echo=TRUE}
model1 <- lm(mpg~am, mtcars)
model2 <- lm(mpg~wt+qsec+cyl+am+hp, mtcars)
model3 <- lm(mpg~., mtcars)

anova(model1, model2, model3)
summary(model2)
```

Conclusion: This model explains 85.8% of the variance. It may be concluded that on average, manual transmissions have 2.729 more mpg than automatic.

## Appendix

``` {r, echo=FALSE}
mtcars$am <- factor(mtcars$am,labels=c('Automatic','Manual'))
```

#### Appendix 1: Histogram of mpg's

``` {r, echo=TRUE, fig.width=5, fig_height=2}
hist(mtcars$mpg, breaks=10, xlab="MPG", main="Histogram of MPGs's")
```

#### Appendix 2: Boxplot of mpg vs. transmission tpye

``` {r, echo=TRUE, fig.width=5, fig.height=3}
suppressWarnings(library(ggplot2))
ggplot(mtcars, aes(x=factor(am), y=mpg, fill=factor(am))) +
  geom_boxplot() + labs(x='Transmission', y='MPG', title='MPG vs Transmission') +
  theme(plot.title = element_text(hjust = 0.5))
```

#### Appendix 3: Scatterplot Matrix of all Variables

``` {r, echo=TRUE, fig.width=8, fig.height=10}
pairs(mpg~., data=mtcars, main='Relationships Between all the Variables')
```

#### Appendix 4: Residual Plots

``` {r, echo=TRUE, fig.width=8, fig.height=8}
par(mfrow=c(2,2))
plot(model2)
```
