---
title: "Practical Machine Learning Project"
author: "Mark Evans"
output: html_document
---

### The Assignment

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

### Data Exploration

Load all necessary packages.

```{r}
options(warn=-1)  #### Turns off warnings
library(caret)
suppressPackageStartupMessages(library(randomForest))
library(rpart)
library(rpart.plot)
library(e1071)
options(warn=0)

set.seed(1234)
```

#### Getting and Cleaning Data

The training and test sets can be found at the following urls:

``` {r}
train_url <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test_url <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'
```

Load the data sets into R. (NA values in the data set are represented by "NA", "#DIV/0!" or "")

``` {r}
training <- read.csv(url(train_url), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(test_url), na.strings=c("NA","#DIV/0!",""))
```

Remove columns in both training and test data sets that have all empty values

``` {r}
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]
```

There are several columns that are not relevant to the outcome predictions: 

- X (id number)
- user_name 
- raw_timestamp_part_1 
- raw_timestamp_part_2 
- cvtd_timestamp 
- new_window 
- num_window 

Remove the columns in the data sets

``` {r}
training <- training[,-c(1:7)]
testing <- testing[,-c(1:7)]
```

### Modeling the data

#### Approach

Two models will be tested using decision tree and random forest. The model with the highest accuracy will be chosen as our final model.

#### Cross-Validation Method

Cross-validation will be performed using createDataPartition() funtion on our training data set randomly without replacement into a 70%/30% split.

#### Expected Out of Sample Error

The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.

``` {r}
split <- createDataPartition(y=training$classe, p=0.7, list=FALSE)
TrainSet <- training[split, ] 
TestSet <- training[-split, ]
```

#### Model #1: Decision Tree

``` {r}
model1 <- rpart(classe ~ ., data=TrainSet)
prediction1 <- predict(model1, TestSet, type = "class")

# Plot the Decision Tree
rpart.plot(model1, main="Basic Decision Tree", extra=102, under=TRUE, faclen=0)
```

Print the confusion matrix for the model. 

``` {r}
confusionMatrix(prediction1, TestSet$classe)
```

#### Model #2: Random Forest

``` {r}
model2 <- randomForest(classe ~. , data=TrainSet)
prediction2 <- predict(model2, TestSet, type = "class")
```

Print the confusion matrix for the model. 

``` {r, echo=TRUE}
confusionMatrix(prediction2, TestSet$classe)
```

The Random Forest model gives us a higher accuracy score (.9969 > .6879). We will use the Random Forest model for our final predictions.

### Predicting Final Outcomes

``` {r, echo=TRUE}
finalpredictions <- predict(model2, testing, type='class')
finalpredictions
```
