---
title: "Statistical Inference Project Part 2"
author: "Mark Evans"
output: pdf_document
---

## Overview

1. Load the ToothGrowth data and perform some basic exploratory data analyses. 
2. Provide a basic summary of the data. 
3. Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose. (Only use the techniques from class, even if there's other approaches worth considering) 
4. State your conclusions and the assumptions needed for your conclusions. 

### Load Datasets

``` {r, echo=TRUE}

library(datasets)
data(ToothGrowth)
tg <- ToothGrowth

```

### Basic Summary of Data

``` {r, echo=TRUE}

summary(tg)

tg$dose <- as.factor(tg$dose)
summary(tg)

table(tg$supp, tg$dose)

## Mean and Standard Deviation of the lengths
c(mean(tg$len, na.rm=TRUE), sd(tg$len, na.rm=TRUE))

## Create box plots of dose vs length faceted by supplement type
suppressWarnings(library(ggplot2))
ggplot(tg, aes(factor(dose), y=len, fill=factor(dose))) +
  geom_boxplot() +
  facet_grid(.~supp) +
  labs(x='Dosage (Milligrams)', y='Tooth Growth') +
  ggtitle('Effect of Dosage and Supplement Types on Tooth Growth')
  
```

### Use confidence intervals and/or hypothesis tests to compare tooth growth by supp and dose.

``` {r, echo=TRUE}

## Subset the tooth data into dosage groups
tg05 <- subset(tg, tg$dose==.5)
tg10 <- subset(tg, tg$dose==1)
tg20 <- subset(tg, tg$dose==2)

## Calculate t.test of tg05 subset
t.test(len ~ supp, data = tg05)

## Calculate t.test of tg10 subset
t.test(len ~ supp, data = tg10)

## Calculate t.test of tg20 subset
t.test(len ~ supp, data = tg20)

```

### State your conclusions and the assumptions needed for your conclusions.

- For dosage .5 and 1, OJ provided the most tooth growth.
- For dosage 2, there is no significant difference between OJ than VC. 
