---
title: "Statistical Inference Project Part 1"
author: Mark Evans
output: pdf_document 
---

## Overview

I will investigate the exponential distribution in R and compare it with the Central Limit Theorem. I will investigate the distribution of averages of 40 exponentials and a thousand simulations.

### Gather exponential distribution and the means of the simulations. 

``` {r, echo=TRUE}

n = 40
s = 1000
lam = 0.2
mns = NULL

for (i in 1:s) {
  
  mns = c(mns, mean(rexp(40, lam)))
  
}

head(mns)
```

### Sample Mean versus Theoretical Mean

``` {r, echo=TRUE}

print(sprintf('Theoretical Mean is %f', 1/lam))

print(sprintf('Sample Mean of the distribution is %f', mean(mns)))

```

### Sample Variance versus Theoretical Variance

``` {r, echo=TRUE}

print(sprintf('Theoretical Variance is %f', (1/lam)^2))

print(sprintf('Sample Variance of the distribution is %f', var(mns)*n))

```

### Distribution

``` {r, echo=TRUE}
suppressWarnings(library(ggplot2))

ggplot(data.frame(x=mns), aes(x=x)) +
  geom_histogram(position='identity', fill='blue', colour='black', binwidth=0.5, aes(y=..density..)) + 
  stat_function(fun=dnorm, colour='red', args=list(mean=5)) +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7, 8, 9), limits=c(1, 9)) +
  scale_y_continuous(breaks=c()) +
  labs(x = "Sample Mean", y = "Frequency", title="Distribution of Samle Means vs Normal Distribution")

```
